Monster or Lambda?  A likelihood analysis

### 2024-07-29 inst/monster_vs_lambda.Rmd
more detail

---
title: "Monster or Lambda?  A likelihood analysis"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

@@

To cite the `hyper2` package in publications, please use @hankin2017_rmd;
takes about an hour to process without cache.  Here I consider
pairwise comparisons with a home-ground advantage and wish to compare
two likelihood functions.  First, the monster parametrization:

@@

## Start simple: two teams

```{r,echo=FALSE}
jj <- matrix(c(NA, 6+2i ,6+5i, NA),2,2,byrow=TRUE)
teams <- LETTERS[1:2]
dimnames(jj) <- list("@home" = teams,"@away"=teams)
dimnames(jj) <- list("@home" = teams,"@away"=teams)
print(jj)
```

Above we see that A played at home 3 times, winning 1 and losing 2
[row 1 column 2 is $1+2i$]; B played at home 9 times, winning 4 and
losing 5 [row 2 column 1 is $4+5i$].  In its most general form, the
$2\times 2$ result matrix would be

@@

and the log-likelihood function would be, algebraically,

@@

Just to check, we evaluate the Hessian matrix and find

@@

@@

This is equal to $(\alpha+\delta)\log\lambda +(\alpha+\gamma)\log
A+(\beta+\delta)\log B -(\alpha+\beta)\log(\lambda A+B)
-(\gamma+\delta)\log(A+\lambda B)$ and even after simplifying [we may
assume $A+B=1$] the result does not appear to have a simple analytical
solution; we must proceed numerically.  To do so,

we need to define a
slight generalization of `home_away3()`:

>>>>

---
title: "Monster or Lambda?  A likelihood analysis"
subtitle: "Takes about an hour to run without cache"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

@@

To cite the `hyper2` package in publications, please use
@hankin2017_rmd.  Here I consider pairwise comparisons with a
home-ground advantage and wish to compare two likelihood functions.
Draws are excluded.  First, the monster parametrization:

@@

# Start simple: two teams {-}

### Monster parametrization {-}

```{r,echo=FALSE}
jj <- matrix(c(NA, 7+2i ,6+5i, NA),2,2,byrow=TRUE)
teams <- LETTERS[1:2]
dimnames(jj) <- list("@home" = teams,"@away"=teams)
dimnames(jj) <- list("@home" = teams,"@away"=teams)
print(jj)
```

Above we see that A played at home 9 times, winning 7 and losing 2
[row 1 column 2 is $7+2i$]; B played at home 11 times, winning 6 and
losing 5 [row 2 column 1 is $6+5i$] (observe that the real components
correspond to home wins and the imaginary components correspond to
away wins.  Thus A wins 7 times at home and 5 times away).  In its
most general form, the $2\times 2$ result matrix would be

@@

and the monster log-likelihood function would be, algebraically,

@@

Observe that the point
$\left(\hat{A},\hat{B},\hat{M}=1-\hat{A}-\hat{B}\right)$ might be
inadmissable.  For example, if $\beta,\delta>0$ and $\alpha=\gamma=0$
then $\hat{A}=\hat{B}=1$ and $\hat{M}=-1$.  This would correspond to
both teams winning all their games at home and losing none.

Just to check the well-definedness of the ML estimate, we evaluate the
Hessian matrix and find

@@

### Lambda parametrization {-}

@@

This is equal to $(\alpha+\delta)\log\lambda +(\alpha+\gamma)\log
A+(\beta+\delta)\log B -(\alpha+\beta)\log(\lambda A+B)
-(\gamma+\delta)\log(A+\lambda B)$ and even after simplifying [we may
assume $B=1-A$] the result of maximizing $\mathcal{S}$ with respect to
$A$ and $\lambda$ does not appear to have a simple analytical
solution; we must proceed numerically.  To do so, we need to define a
slight generalization of `home_away3()`:
